SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data

### 2011-09-29 seqbias-paper.tex
notes on non-empty probability

These biases may adversely effect transcript discovery, as low level noise may
be over-reported in some regions, and in others, active transcription may be
under-reported. They render untrustworthy comparisons of relative abundance
between genes or isoforms, and any test of differential expression hangs on the
assumption that these biases are identical between replicates.  Additionally, in
many tests of differential expression---similar to the effects of transcript
length---a higher read count will result in higher statistical confidence. It
follows that the sensitivity of such a test will also be biased by sequence,
affecting downstream analysis such as gene ontology enrichment tests.

@@

\citet{Li2010} propose two models. The first is a Poisson linear
model, in which read counts across a transcript follow an inhomogeneous Poisson
process. The read count at position $i$ within the transcript is Poisson
distributed with parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of
weights determined by the nucleotide at each position surrounding the read
start, in addition to a term capturing the abundance of the transcript.

@@

Another model, proposed by \citet{Hansen2010}, directly estimates the
distribution of initial heptamers within reads, then estimates a presumed
background heptamer distribution, sampled from the ends of reads. The read count
at a given position is then adjusted by the ratio of the foreground and
background heptamer probabilities. Specifying two distributions over heptamers
(i.e. foreground and background distributions) requires \mbox{$2(4^7-1) =
32,766$}
parameters, so while no gene annotations are needed to train such a model, a
significant number of reads are required.

@@

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the model is evaluated
at each position within a locus, predicting bias. The predicted bias can then be
used to adjust read counts, as in (e). In (d) and (e) we show the results of
this method applied to the $3'$ UTR of Apoa2, using data from
\citet{Mortazavi2008}.}
\label{fig:overview}
\end{figure}

@@

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{models.eps}}
\caption{
The network structures learned on each of the data sets are displayed. Positions
are relative to the read start, which is labeled 0. Hollow circles indicate
positions that were not included in the model, being deemed uninformative, given
the other positions and edges. The number of parameters needed to
specify each model is listed in parenthesis below. Applied to data with less
bias, a sparser model is trained, as evinced by the Wetterbom data set.
}
\label{fig:models}
\end{figure*}

@@

@@

If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
frequency, the KL divergence is computed as
$$D_k( f, f' ) = \sum_{x} \left( f_x \log_{2}( f_x / f'_x ) + f'_x \log_{2}( f'_x / f_x) \right)$$
where the sum is over all $k$-mers.

@@

This measure can be interpreted as the improvement in fit over the null model,
with $R^2 = 1$ indicating a prefect fit, and smaller numbers, an increasingly
worse fit. A more dubious, yet intuitive interpretation of $R^2$ is the
proportion of variability explained by the model.

>>>>

These biases may adversely effect transcript discovery, as low level noise may
be over-reported in some regions, and in others, active transcription may be
under-reported. They render untrustworthy comparisons of relative abundance
between genes or isoforms, and any test of differential expression hangs on the
assumption that these biases are identical between replicates, an undesirable
assumption given that the causes of the bias are not well understood.
Additionally, in many tests of differential expression---similar to the effects
of transcript length---a higher read count will result in higher statistical
confidence. It follows that the sensitivity of such a test will also be biased
by sequence, affecting downstream analysis such as gene ontology enrichment
tests.

@@

\citet{Li2010} propose two models. The first is a Poisson linear model, in which
read counts across a transcript follow an inhomogeneous Poisson process. The
read count at position $i$ within the transcript is Poisson distributed with
parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of independent
weights determined by the nucleotide at each position surrounding the read
start, in addition to a term capturing the abundance of the transcript.

@@

Another model, proposed by \citet{Hansen2010}, directly estimates the
distribution of initial heptamers within reads, then estimates a presumed
background heptamer distribution, sampled from the ends of reads. The read count
at a given position is then adjusted by the ratio of the foreground and
background heptamer probabilities. Specifying two distributions over heptamers
(i.e. foreground and background distributions) requires \mbox{$2(4^7-1) =
32,766$} parameters, so while no gene annotations are needed to train such a
model, a significant number of reads are required, and a number that increases
exponentially with $k$, if it were desirable to to model $k$-mers for $k > 7$.

@@

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the model is evaluated
at each position within a locus, predicting bias. The predicted bias can then be
used to adjust read counts, as in (e). In (d) and (e) we show the results of
this method applied to the $3'$ UTR of Apoa2, using data from
\citet{Mortazavi2008}. In bias coefficients predicted across 10 million positions
of chromosome 1, the log10 bias of 95\% of the positions were between -1.14 and
0.63, suggesting that most adjustments are not large.
}
\label{fig:overview}
\end{figure}

@@

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{models.eps}}
\caption{
The network structures learned on each of the data sets are displayed. Positions
are relative to the read start, which is labeled 0. Hollow circles indicate
positions that were not included in the model, being deemed uninformative, given
the other positions and edges. The number of parameters needed to
specify each model is listed in parenthesis below. Applied to data with less
bias, a sparser model is trained, as evinced by the Wetterbom data set. Note
that dependencies (i.e., arrows) tend to span a short distances, and nodes tend
to have a small in-degree (i.e., have few inward arrows). In practice, we save
time in training by prohibiting very distant dependencies (greater than 10, by
default) or very high in-degrees (greater than 4, by default).
}
\label{fig:models}
\end{figure*}

@@

Evident from Figure \ref{fig:overview}, the assumption of uniform read coverage
often does not hold in typical RNA-Seq data sets. Although the bias corrected
read counts across the exon pictured in this example are visibly more uniform,
we sought a simple, objective tests that could be applied genome-wide. To this
end, we used cross-validation tests (i.e., methods were trained and tested on
disjoint subsets of the same RNA-Seq data sets) of a quantitative measure of the
increase in uniformity of nucleotide frequencies (Kullback-Leibler divergence in
Section 3.1) and increase in uniformity of read coverage (Poisson regression in
Section 3.2). Additionally, we compare RNA-seq based estimate of gene expression
to qRT-PCR based estimated for the same genes, showing increased correlation
between the two methods after bias correction (Section 3.3).

@@

If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
frequency, the KL divergence is computed as
$$D_k( f, f' ) = \sum_{x} \left( f_x \log_{2}( f_x / f'_x ) + f'_x \log_{2}( f'_x / f_x) \right)$$
where the sum is over all $k$-mers. This can be thought of a measure
dissimilarity between two probability distributions. If $f_x$ and $f'_x$ for a
$k$-mer $x$ are approximately equal, their log-ratio will also be approximately
zero, leading to a small KL divergence (zero, when the distributions are equal).
Conversely, very different $k$-mer frequencies will result in a larger KL
divergence.

@@

This measure can be interpreted as the improvement in fit over the null model,
with $R^2 = 1$ indicating a prefect fit, occurring model being evaluated
achieves a likelihood of 1. Smaller number indicate an increasingly worse fit,
with $R^2 = 0$ representing no improvement over the null model, and $R^2 = 0.5$,
for example, indicating the model has a log-likelihood equal to half that of the
null model (a large improvement, as log-likelihoods negative or zero). This
measure can also take on values less than 0, indicating that the model has worse
fit than the null model (i.e., when adjusting read counts according the bias
coefficients leads to less uniform read coverage).

### 2011-09-29 seqbias-supplement.tex
notes on non-empty probability

>>>>

\section{TODO: title}

In the case that an experiment is unbiased, we would like an empty model to be
trained, so that using the method to correct for bias will have no effect. Any
non-empty model would otherwise decrease the accuracy of the quantification. To
address this concern, we derive here an upper bound on the probability of a
non-empty model being trained, when the data is unbiased.

The training procedure begins by considering the the set of models formed by
including each single nucleotide position from within the training sequences.
For each single nucleotide model, the Bayesian infromation criterion is evaluated. If in any of these
model the BIC score increases over the empty model, a non-empty model will be
trained, otherwise the training procedure will halt with an empty model.

We begin by considering one single nucleotide model. Suppose the background and
foreground nucleotide distributions for this single position are equal (i.e.,
the experiment is unbiased in this position) specified by the parameters $(p_{a}, p_{c},
p_{g}, p_{t})$.

The model is trained on a set of $n$ foreground and $n$ background sequences.
Let $X = (x_{a}, x_{c}, x_{g}, x_{t})$ give the number of times each nucleotide
was observed in the foreground sample, and $Y = (y_{a}, y_{c}, y_{g}, y_{t})$
in the background sample.

First, suppose the nucleotide distribution estimated from $X$ and from $Y$ both
have KL divergence bounded by some number $\epsilon$:
$$ D_{X} = \sum_{k \in \{a, c, g, t\}} (x_k / n) \log \left( \frac{x_k / n}{p_k}
\right) < \epsilon $$
and
$$ D_{Y} = \sum_{k \in \{a, c, g, t\}} (y_k / n) \log \left( \frac{y_k / n}{p_k}
\right) < \epsilon $$

Now, objective function evaluated is the Bayesian information criterion (BIC)
applied to the conditional log-likelihood (CLL).
$$ \text{BIC} = 2 \cdot \text{CLL} - m \log n $$
where CLL in this single nucleotide case can be written as,
$$ \text{CLL} =
\sum_{k \in \{a, c, g, t\}}
\left(
x_k \log \left( \frac{x_k/n}{(x_k + y_k)/n} \right) +
y_k \log \left( \frac{y_k/n}{(x_k + y_k)/n} \right) \right)$$

The KL divergence obtained by combining the two samples $X$ and $Y$ is,
$$D_{XY} = \sum_{k \in \{a, c, g, t\}}
\frac{(x_k + y_k)}{2n}
\log \left( \frac{(x_k + y_k)/2n}{p_k} \right)$$

The KL divergence is non-negative, and so adding $2n D_{XY}$ to the CLL, we get
\begin{align*}
CLL &\le CLL + 2n D_{XY} \\
&=
\textstyle{
\sum_{k \in \{a, c, g, t\}} 
x_k \log \left( \frac{x_k/n}{(x_k + y_k)/n} \right)
+ y_k \log \left( \frac{y_k/n}{(x_k + y_k)/n} \right)
+ (x_k + y_k) \log \left( \frac{(x_k + y_k)/2n}{p_i} \right)} \\
&=
\textstyle{
\sum_{k \in \{a, c, g, t\}}
\left(
x_k \log \left( \frac{x_k/n}{2 p_i} \right) +
y_k \log \left( \frac{y_k/n}{2 p_i} \right)
\right)} \\
&=
\textstyle{
n D_{X} + n D_{Y} + \sum_{k \ in \{a, c, g, t\}} (x_k + y_k) \log (1/2)
} \\
%\textstyle{
%n \sum_{k \in \{a, c, g, t\}}
%\left( (x_k / n) \log \left( \frac{x_k/n}{p_i} \right) \right) +
%n \sum_{k \in \{a, c, g, t\}}
%\left( (y_k / n) \log \left( \frac{y_k/n}{p_i} \right) \right) +
%\sum_{k \in \{a, c, g, t\}} (x_k + y_k) \log (1/2)} \\
&< 2n\epsilon + 2n \log (1/2)
\end{align*}
Thus, if the KL divergence of the two samples $X$ and $Y$ are bounded, the
conditional log-likelihood is as well.

The BIC score of an empty model is,
$$\text{BIC}_0 = 2 \sum_{k \in \{a, c, g, t\}}
\left(
x_k \log(1/2) + y_k \log(1/2) \right) - m
\log n = 4 n \log (1/2) $$
where the number of parameters needed to specify the empty model is $m = 0$.

The BIC score of the single nucleotide model, when the KL divergence of $X$ an
$Y$ is bounded by $\epsilon$ is
\begin{align*}
\text{BIC} &= 2 \cdot CLL - m \log n \\
&= 2 \cdot CLL - 6 \log n \\
&< 4n \epsilon + 4 n \log (1/2) - 6 \log n
\end{align*}

It follows that if
\begin{align*}
4n \epsilon + 4 n \log 1 / 2 - 6 \log n  &\le   4 n \log 1/2 \\
\epsilon &\le \frac{3}{2n} \log n
\end{align*}
then
$$ \text{BIC} \le \text{BIC}_0 $$

In summary, if $D_X, D_Y < \frac{3}{2n} \log n$, then $\text{BIC} \le
\text{BIC}_0$ and the empty model is retained. That is, the probability of an
empty model being trained is bounded below by the probability that the KL
divergence of both sampled is bounded above $\frac{3}{2n} \log n$:
$$ \Pr(D_{X} < \frac{3}{2n} \log n) \cdot 
\Pr(D_{Y} < \frac{3}{2n} \log n) $$

Since $X$ and $Y$ are drawn from the same distribution, $D_{X}$ and $D_{Y}$ are
identically distributed as well, and so we can simplify this formula as,
$$ (\Pr(D < \frac{3}{2n} \log n))^2 $$
where $D$ is the divergence of any sample of size $n$. The probability of a
non-empty model being trained is then less than
$$ 1 - (\Pr(D < \frac{3}{2n} \log n))^2 $$

In training we consider mulitple nucleotides. If $h$ nucleotides are considered,
the probability of a non-empty model being trained is less than
$$ 1 - (\Pr(D < \frac{3}{2n} \log n))^{2h} $$

As discussed in Supplementary Section 9.2, $2n D$ is approximated by a $chi^2$
distribution with 3 degrees of freedom for large $n$, and so we can estimate
this probability for various values of $n$. This is shown in Figure
\ref{fig:nepr}. From this figure we can see, for example that training with more
than 10,000 reads results in an incorrect non-empty model with probability less
than 0.0004.

In the low end, the upper bound on non-empty model probability is much higher.
In our tests, this upper bound is not tight for a small number of reads. We
trained the model on 100 sets of 100 simulated reads, drawn randomly at uniform
from exonic sequences and in only 3 trials was a non-empty model trained, far
less than the upper bound prediction of 23 out of 100. This test was repeated
using sets of 50,000 reads, and in no trial was a non-empty model trained.

From these results, and the results in Supplementary Section 5, we recommend
training our method with at least 10,000 reads.

\begin{figure}[H]
\centerline{
\includegraphics[width=0.8\textwidth]{fig/nepr.pdf}}
\caption{Plotted is the upper bound on the probability of training a non-empty
model when there is no real bias in the experiment when the model is trained
with various number of reads $n$. We fixed the number of nucleotides being
considered as $h = 41$, as this what was used for the evaluation in Section 3.}
\ref{fig:nepr}
\end{figure}
